---
title: "Editable tables in shiny"
subtitle: "Week4 pitch presentation for Coursera's Developing Data Products"
author: "Daniel Bader"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: slidy_presentation
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message = F, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plotly)
library(rhandsontable)
```


## Motivation

Create a shiny template for use with my actual work data 
that has the following features:

- Data with numeric and categorical variables
- Text search to subset the data
- Editable table powered by <code>rhandsontable</code>
- Interactive plot based on subsetted and potentially modified table
- Export of standalone HTML report to a file using the modified data


## Data: <code>mtcars</code>

Quarter mile speed in seconds vs horse power

```{r cars, echo = TRUE}
plot_ly(data = mtcars, x = ~hp, y = ~qsec, hovertext = rownames(mtcars)) %>%
    add_markers()
```

## Editable table

```{r}
rhandsontable(data = mtcars, search = T) %>%
    hot_table(highlightRow = TRUE, rowHeaderWidth = 200) %>%
    hot_cols(columnSorting = TRUE)
```


## Not enough?

https://baderd.shinyapps.io/Dynamic_mtcars/ 
